Balsa: A Compact Line Notation Based on SMILES

### 2022-09-09 manuscript.tex
Revises manuscript line-by-line.

Despite the central role that SMILES plays in cheminformatics, public descriptions of the language remain imprecise at best. A 1987 report to the US Environmental Protection Agency by Weininger described a language similar to SMILES, but with important differences~\cite{anderson:1987}. Weininger's 1988 paper offered an overview of the language, but omitted or only partially addressed points crucial to correctly implementing SMILES, including: (1) stereochemical configuration; (2) double bond conformation; (3) the encoding and decoding of \enquote{aromatic} features; (4) the computation of implicit hydrogen count, especially in the context of aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2003}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. (\enquote{Daylight}), recapitulates previously-published material~\cite{daylightTheory}. The lack of other authoritative sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}. OpenSMILES was conceived as \enquote{a non-proprietary specification for the SMILES language.} Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, later adapted to a standard parser toolchain~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of \enquote{standard form.} Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

@@

The lack of a comprehensive SMILES specification has caused several immediate problems. Maintainers of SMILES implementations lack a blueprint for working toward a common set of features and behaviors. At the same time, authors of new SMILES implementations must address ambiguities on an ad hoc basis. The diversity of implementations combined with language underspecification leaves standards bodies with a limited pool of source material upon which to draw. Widespread divergence around the edges of the SMILES language hampers the development of compliance suites, which only amplifies the problems faced by software teams trying to build compliant software. Finally, extensions to SMILES only make sense in the context of a rigorously defined base language.

More broadly, this situation undermines ongoing data integrity efforts underway in other scientific disciplines. For example, the FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. The lack of a well-developed standard means that chemical data sets based on SMILES will suffer in all four areas.

The very notion of a SMILES standard may not be realistic at this point. Today a dozen or more slightly different implementations are in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language drift while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of the documentation that does exist. The Daylight implementation, like some of the others, is distributed under a commercial license that restricts use~\cite{daylightToolkit}, and a publicly-accessible web service operated for many years has been decommissioned~\cite{depictArchive}. Although SMILES may underpin a lot of modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Balsa's main goal is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This allows general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (\enquote{molfile})~\cite{ctfileFormats}.

A secondary goal is brevity. Molecules should be expressable using only a few characters for each atom and either zero or one character for each bond. Brevity benefits humans who manually encode chemical structures within interactive line-driven interfaces such as those found in terminal sessions and notebooks. Brevity also benefits computers, for which compact molecular representation can improve the performance when storage, bandwidth, or processor cycles are limiting factors. Figure~\ref{fig:representative-examples} presents some simple chemical structures together with their Balsa serializations.

@@

Balsa's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}. Syntax defines the set of valid strings for a language, whereas semantics determine the meaning these strings. Balsa semantic are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

@@

A related but more common concept is the \textit{molecular graph}~\cite{balaban:1985}. Unlike molecule graphs, molecular trees do not support cycles or disconnected components. This difference might appear to make molecular trees unsuitable for molecular representation, where both rings and aggregation play important structural roles. However, these apparent shortcomings can be mitigated and even harnessed.

The main advantage of trees over graphs is information density. Graphs must accommodate cycles, which is accomplished by assigning a unique identifier to each node. An edge references two of these identifiers. Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges. In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges. This simplifies the representation and eliminates the overhead of assigning and using node identifiers.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{molecular-graph-vs-tree.pdf}
    \caption{Molecular Graph vs. Molecular Tree. The bonds in a molecular graph reference atoms indirectly through unique indexes (lower left). In contrast, a bond in a molecular tree defines its target atom inline (lower right). This increases the potential efficiency of trees over graphs for molecular representation. To simulate ring closure, a molecular tree introduces bridge nodes, as described later (solid grey circles).}
    \label{fig:molecular-graph-vs-tree}
\end{figure}

When discussing molecular trees, it's often useful to refer to \textit{node order} (Figure~\ref{fig:node-order}). An order is a property of a set that allows its members to be sorted through pairwise comparisons. Node order is the order in which the nodes of a molecular tree are added through depth-first construction. A node added before another node is said to \enquote{precede} it. A node added after another node is said to \enquote{succeed} it. In this way nodes can be prioritized. The root atom, uniquely lacking a parent, has highest priority because it precedes every other node. Similarly, a parent always precedes its children. Finally, the first child added to a given parent precedes all of its siblings. This relationship is transitive. If a node precedes (or succeeds) a sibling, then so do all of its descendants.

@@

Cycles are supported through \textit{bridges}. A bridge consists of two bond-node pairs that together represent one bond. One ach side of a bridge and atom connects to a bond, but the two atoms are not connected to each other. Instead, each atom is succeeded by a \textit{bridge node}. Unlike an atom, a bridge node has no children and so may not serve as the root of the tree. A bridge preserves the immediate atomic environment around each atom, allowing many of the same atomic computations possible in molecular graphs. Any bond, whether it closes a ring or not, can be represented as a bridge.

Each bridge node serves as a proxy to a \textit{referenced atom}. Should a bridge node be present during an operation that requires an atom, the referenced is used instead. Although a bridge node always succeeds its parent, the referenced atom may precede or succeed that parent.

Multiple molecules are supported through \textit{gaps}. A gap is an edge devoid of bonding. Gaps can be used whenever two or more molecules need to be encoded within the same molecular tree. Specific applications include ionic bonding and non-ionic complexes. However, no bonding relationship whatsoever is implied by the presence of a gap. In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.

@@

Gaps and bridges decouple connectivity from bonding, but for opposite purposes. Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them. These relationships can complicate the analysis of connectivity and bonding. For example, the atom succeeding a gap behaves in most situations as if it were a root. When tallying the bonds surrounding such an atom, only children are considered. Similar considerations apply when a gap succeeds an atom. Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond. Nevertheless, a bridge node merely stands in for another atom found elsewhere in the tree. Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.

@@

Within a molecular tree, atoms and bonds play complementary roles. Atoms control nuclear particle counts, valence electron counts, and configuration. Bonds determine bonding electron counts and conformation. Constitution, configuration, and conformation are all influenced by the individual and collective behavior of atoms and bonds.

At the lowest level, atom and bond behavior are governed by their respective \textit{attributes}. An attribute is a key/value pair in which the key is a name composed of a character sequence and the value is constrained by a single, immutable \textit{type}. Types can be composed and decomposed, a feature supported by many programming languages.

@@

The attributes \ttt{element} (\enquote{element}) and \ttt{isotope} (\enquote{isotope}) define atomic number and neutron count, respectively. Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:periodic-table}). An atom whose element is unknown must have an \ttt{element} attribute equal to \ttt{None}. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, defined as the sum of proton and neutron count. Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance. Otherwise, the lower bound of the value of \ttt{isotope} is one. This lower limit allows for physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5). Implementations may reject such atomic states as invalid.

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen atom and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible to do so (see: \nameref{computing-implicit-hydrogen-count}).

@@

\begin{enumerate}
    \item{If implicit hydrogen counting is used, default values of isotope, charge, and parity must also be used. Implicit hydrogen counting is only supported for elements in the list: B; C, N; O; P; S; F; Cl; Br; or I.}
    \item{A selected atom must use one of the following five elements: B; C; N; O; P; or S.}
\end{enumerate}

A bond is composed of four attributes (Table~\ref{table:bond-attributes}). The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two. A bond's order may only assume the values one, two, or three. The \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization and participate in extended bonding. The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond. The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down}, will be described in detail later (see: \nameref{conformation}). The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).

@@

@@

The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics. Implementations may, of course, use any suitable internal data model. They must, however, ensure that any internal data model is consistent with the one provided here.

@@

Electron counting in Balsa is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even number of electrons. These electrons are drawn in equal quantities from each atom's set of valence electrons. A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total). The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero. Bonding to such an atom adds the required number of electrons directly to the bond.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{electron-counting.pdf}
    \caption{Electron Counting. A root atom and its child are added to a tree (left). Next, a bond is added (center). Finally, the appropriate electron count is deducted from each atom and added to the bond (right).}
    \label{fig:electron-counting}
\end{figure}

@@

Membership of a DS is determined as follows. An atom is added to the DS by setting its \ttt{selected} attribute to \ttt{true} through a process called \textit{selection}. As noted previously (see: \nameref{atoms-and-bonds}), only some atomic states are compatible with selection. An edge will be added to the DS if both of the atoms it spans are selected and the bond itself is elided. This allows arbitrary bonds between selected atoms to be excluded from the DS. If a bond's target is a bridge node, the corresponding edge will be added to the DS if both the parent atom and the equivalent target atom are selected.

@@

Given an atom whose element has one or more default valences, \textit{subvalence} can be computed (Algorithm~\ref{algorithm:subvalence}). Subvalence is the number of hydrogens that can be added to an atom without exceeding the lowest possible default valence. If no suitable default valence exists, then subvalence equals zero. This computation can be expressed algorithmically (Algorithm~\ref{algorithm:subvalence}). First, the atom's valence is computed. Next, the ordered list of default valences of the atom's element is found. For each default valence, the difference between it and the valence is compute. If this differences is positive, it is returned as the subvalence. If no suitable default valence is found, zero is returned. The subvalence of an atom whose element has no default valences (i.e., no listing in Table~\ref{table:default-valences}) is zero. For example, an iron atom always has a subvalence of zero, regardless of how it is bonded.

@@

For a representative subvalence computation, consider an isolated nitrogen atom without virtual hydrogens. The atom's bond order sum is zero. Its default valences are 3 and 5. The difference is found to be three ($3 - 0$). Therefore, the subvalence of this atom is three.

@@

The phosphorous atom in phosphorous acid (H\textsubscript{3}PO\textsubscript{3}) illustrates the use of Algorithm \ref{algorithm:subvalence} for atoms with multiple target valences. Bond order sum is four ($2 + 1 + 1$). The first default valence is 3, but subtracting valence from that value yields a negative number ($3 - 4$). Continuing to the next default valence, five, a difference of 1 is obtained. Therefore, the subvalence of the phosphorous-bearing atom is reported as one.

@@

If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported. The meaning of such an atomic state may seem suspect because an atom without a free valence can not perform the required promotion of a single bond during deselection. As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition. Returning zero avoids miscalculation of the implicit hydrogen count.

An atom's implicit hydrogen count may or may not correlate with chemical intuition or experimental data. Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}). We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2). However, the subvalence for the phosphorous atom is found to be three ($2 + 1 - 3$). The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two. To represent this atom and other like it, virtual hydrogens must be used.

@@

The same procedure works in reverse when encoding the \ttt{direction} attribute (Figure~\ref{fig:assigning-ppb}). 2-butene having the \textit{anti} conformation is modeled using a diagram whose left- and right-hand sides point downward and upward, respectively. The parent atom, a primary carbon, is identified. It is bonded to one neighbor through a PPB. To match the diagram, a direction of \ttt{Up} is assigned. The child atom is attached to its own child through a double bond. Finally, the grandchild is attached to its own child through a PPB whose direction is consistent with the diagram. A direction of \ttt{Up} is therefore assigned.

@@

An exception to the underspecification rule applies in the case of conjugated polyenes (Figure~\ref{fig:bond-error-state-exceptions}). Here, a PPB may run from parent to child but not from grandchild to great-grandchild. No error is reported in such a case and the conformation of the terminal double bond remains undefined.

@@

Cyclic conjugated polyenes present special challenges to the use of PPBs, as exemplified by cyclooctatetraene (Figure~\ref{fig:cyclooctatetraene}). Although the (\textit{anti}, \textit{anti}, \textit{anti}, \textit{anti}) conformation is expressable, the (\textit{syn}, \textit{anti}, \textit{anti}, \textit{anti}) conformation is not. This problem arises because the conformation for two different double bonds is set by the same PPB.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{cyclooctatetraene.pdf}
    \caption{Cyclooctatetraene. The all-(\textit{Z}) conformation is expressable (left), but the (\textit{Z},\textit{Z},\textit{Z},\textit{E}) conformation is not (right). Although the local conformations for the latter are all valid, the global conformation introduces a conflict, the direction marked in red.}
    \label{fig:cyclooctatetraene}
\end{figure}

@@

Balsa limits configuration to the special case of an atom with exactly four substituents placed at the vertices of a tetrahedron. Here, \enquote{substituent} means an atomic neighbor or virtual hydrogen. Lone electron pairs are not considered to be substituents.

The configuration about an atom can specified through its \ttt{parity} attribute. This attribute, together with a general convention, determine the relative three-dimensional positioning of substituents about a central atom (Figure~\ref{fig:atom-parity}). First, the parent of the central atom is identified. Then the frame of reference is shifted by sighting down the bond from the central atom's parent to the central atom. An equilateral triangle is then inscribed about the central atom. The bonds from the central atom are then iterated in order. If the parity of the central atom equals \ttt{Clockwise}, the child atoms attached through each bond are placed around the triangle in clockwise order. Otherwise the children are arranged in counterclockwise order.

@@

One additional modification applies to a central atom without a parent. The frame of reference is shifted to sight along the direction of the first child. The remaining children are then placed around the triangle in either counterclockwise or clockwise order according to the value of the central atom's \ttt{parity} attribute. If a central root atom bears a virtual hydrogen, it is considered to be the first child.

@@

It is sometimes useful to manipulate a configuration in a way that preserves the relative three-dimensional positioning of neighbor atoms. This process is called \textit{transformation}. Four operations suffice to transform any configuration into any other (Figure~\ref{fig:configurational-transformations}):

>>>>

Despite the central role that SMILES plays in cheminformatics, public descriptions of the language remain imprecise. A 1987 report to the US Environmental Protection Agency by Weininger described a language similar to SMILES, but with important differences~\cite{anderson:1987}. Weininger's 1988 paper offered an overview of the language, but omitted or only partially addressed points crucial to correctly implementing SMILES, including: (1) stereochemical configuration; (2) double bond conformation; (3) the encoding and decoding of \enquote{aromatic} features; (4) the computation of implicit hydrogen count, especially in the context of aromatic features; (5) syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2003}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. (\enquote{Daylight}), recapitulates published material~\cite{daylightTheory}. The lack of other authoritative sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}. OpenSMILES was conceived as \enquote{a non-proprietary specification for the SMILES language.} Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, later adapted to a standard parser toolchain~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of \enquote{standard form.} Absent were detailed procedures for composing and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

@@

The lack of a comprehensive SMILES specification has caused several practical problems. Maintainers of SMILES implementations lack a blueprint for working toward a common set of features and behaviors. At the same time, authors of new SMILES implementations must address ambiguities on an ad hoc basis. The diversity of implementations combined with language underspecification leaves standards bodies with a limited pool of source material upon which to draw. 
Divergence at the edges of the SMILES language hampers the development of compliance suites, amplifying the problems faced by software teams trying to build compliant software. Finally, extensions to SMILES only make sense in the context of a rigorously defined base language.

More broadly, this situation undermines ongoing data integrity efforts underway in other scientific disciplines. As an example, the FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. The lack of a precise specification means that chemical data sets based on SMILES will suffer in all four areas.

The very notion of a SMILES standard may not be realistic at this point. Today a dozen or more slightly different implementations are in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language drift while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of the documentation that does exist. The Daylight implementation, like some of the others, is distributed under a commercial license that restricts use~\cite{daylightToolkit}, and a publicly-accessible web service operated for many years was decommissioned~\cite{depictArchive}. Although SMILES may underpin a lot of modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Balsa's main goal is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This allows general-purpose devices and networks to convey complex domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (\enquote{molfile})~\cite{ctfileFormats}. Figure~\ref{fig:representative-examples} presents some simple chemical structures together with their Balsa serializations.

A secondary goal is brevity. Molecules should be expressable using only a few characters for each atom and either zero or one character for each bond. Brevity benefits humans who manually encode chemical structures within interactive line-driven interfaces such as those found in terminal sessions and notebooks. Brevity also benefits computer handling, where the use of compact molecular representation can improve performance when storage, bandwidth, or processor cycles are limiting factors.

@@

Balsa's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}. Syntax defines the set of grammatically valid strings for a language, whereas semantics determines their meaning. Balsa semantic are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

@@

A related but more common concept is the \textit{molecular graph}~\cite{balaban:1985}. Unlike molecule graphs, molecular trees do not support cycles or disconnected components. This difference might appear to make molecular trees unsuitable for molecular representation, in which both rings and aggregation play important structural roles. However, these apparent shortcomings can be mitigated and even harnessed.

The main advantage of trees over graphs is information density. Graphs must accommodate cycles, a requirement fulfilled by assigning a unique identifier to each node. Each edge then references two of these identifiers. Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges. In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges. This simplifies the representation and eliminates the overhead of assigning and using node identifiers.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{molecular-graph-vs-tree.pdf}
    \caption{Molecular Graph vs. Molecular Tree. The bonds in a molecular graph reference atoms indirectly through unique indexes (lower left). In contrast, a bond in a molecular tree defines its target atom inline (lower right). This increases the potential efficiency of trees over graphs for molecular representation. To simulate ring closure, a molecular tree introduces bridge nodes (solid grey circles).}
    \label{fig:molecular-graph-vs-tree}
\end{figure}

When discussing molecular trees, it's often useful to refer to \textit{node order} (Figure~\ref{fig:node-order}). An order is a property of a set that allows its members to be sorted through pairwise comparisons. Node order is the order in which the nodes of a molecular tree are added through depth-first construction. A node added before another node is said to \enquote{precede} it. A node added after another node is said to \enquote{succeed} it. In this way nodes can be prioritized. The root atom, uniquely lacking a parent, is ordered first because it precedes every other node. Similarly, a parent always precedes its children. Finally, the first child added to a given parent precedes all of its siblings. This relationship is transitive. If a node precedes (or succeeds) a sibling, then so do all of its descendants.

@@

Cycles are supported through \textit{bridges}. A bridge consists of two bond-node pairs that together represent one bond. On each side of a bridge an atom connects to a bond, but the two atoms are not connected to each other. Instead, each atom is succeeded by a \textit{bridge node}. Unlike an atom, a bridge node has no children and so may not serve as the root of the tree. A bridge preserves the immediate atomic environment around each atom, allowing many of the same atomic computations possible in molecular graphs. Any bond, whether it closes a ring or not, can be represented as a bridge.

Each bridge node serves as a proxy to a \textit{referenced atom}. Should a bridge node be present during an operation that requires an atom, the referenced atom is used instead. Although a bridge node always succeeds its parent, the referenced atom may precede or succeed that parent.

Multiple molecules are supported through \textit{gaps}. A gap is an edge devoid of bonding. Gaps can be used whenever two or more disconnected molecules need to be encoded within the same molecular tree. Specific applications include ionic bonding and non-ionic complexes. However, no bonding relationship whatsoever is implied by the presence of a gap. In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.

@@

Gaps and bridges decouple connectivity from bonding, but for opposite purposes. Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them. These relationships can complicate the analysis of connectivity and bonding. For example, the atom succeeding a gap behaves in most situations as if it were a root. When tallying the bonds surrounding such an atom, only children are considered. Similar considerations apply when a gap succeeds an atom. Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond. Nevertheless, a bridge node merely stands in for a referenced atom found elsewhere in the tree. Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.

@@

Within a molecular tree, atoms and bonds play complementary roles. Atoms define nuclear particle counts, valence electron counts, and configuration. Bonds define bonding electron counts and conformation. Constitution, configuration, and conformation are all influenced by the individual and collective behavior of atoms and bonds.

At the lowest level, atom and bond behavior are governed by their respective \textit{attributes}. An attribute is a key/value pair in which the key is a name composed of a character sequence and the value is constrained by a single, immutable \textit{type}. Types can be composed, a feature supported by many programming languages.

@@

The attributes \ttt{element} (\enquote{element}) and \ttt{isotope} (\enquote{isotope}) define atomic number and mass number, respectively. Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:periodic-table}). An atom whose element is unknown must have an \ttt{element} attribute equal to \ttt{None}. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, defined as the sum of proton and neutron count. Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance. Otherwise, the lower bound of the value of \ttt{isotope} is one. This lower limit allows for physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5). Implementations may reject such atomic states as invalid.

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen atom and its bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible for this treatment (see: \nameref{computing-implicit-hydrogen-count}).

@@

\begin{enumerate}
    \item{If implicit hydrogen counting is used, default values of isotope, charge, and parity must also be used. Implicit hydrogen counting is only supported for elements in the list: B; C, N; O; P; S; F; Cl; Br; or I.}
    \item{An atom can only be selected if its element attribute is one of: B; C; N; O; P; or S.}
\end{enumerate}

A bond is composed of four attributes (Table~\ref{table:bond-attributes}). The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two. A bond's order may only assume the values one, two, or three. The \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization and participate in extended bonding. The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond. The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down} (Table~\ref{table:bond-direction-variants}), will be described in detail later (see: \nameref{conformation}). The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).

@@

\begin{table}
    \caption{Bond Direction Variants.}
    \centering
    \begin{tabular}{l p{6.6cm}}
        \hline
        Variant & Description \\
        \hline
        \ttt{Up} & The target node lies above the parent. \\
        \tt{Down} & The target node lies below the parent. \\
        \hline
    \end{tabular}
    \label{table:bond-direction-variants}
    \end{table}

@@

The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics. Implementations may, of course, use any suitable internal data model. They must, however, ensure that such a model is consistent with the one provided here.

@@

Electron counting in Balsa is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even electron count. These electrons are drawn in equal quantities from each atom's set of valence electrons. A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total). The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero. Bonding does not change it.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{electron-counting.pdf}
    \caption{Electron Counting. A root atom and its child are added to a tree (left, center left). Next, a bond is added (center right). Finally, the appropriate electron count is deducted from each atom and added to the bond (right).}
    \label{fig:electron-counting}
\end{figure}

@@

Membership of a DS is determined as follows. An atom is added to the DS by setting its \ttt{selected} attribute to \ttt{true} through a process called \textit{selection}. As noted previously (see: \nameref{atoms-and-bonds}), only some atomic states are compatible with selection. A bond is added to the DS if both of the atoms it spans are selected and the bond itself is elided. This allows arbitrary bonds between selected atoms to be excluded from the DS by ensuring they are not elided. A bond to a bridge node will be added to the DS only if both the parent and the referenced atom are selected.

@@

Given an atom whose element has one or more default valences, \textit{subvalence} can be computed (Algorithm~\ref{algorithm:subvalence}). Subvalence is the number of hydrogens that can be added to an atom without exceeding the lowest possible default valence. If no suitable default valence exists, then subvalence equals zero. The algorithm begins by computing the atom's valence. Next, the ordered list of default valences of the atom's element is found. For each default valence, the difference between it and the valence is compute. If this differences is positive, it is returned as the subvalence. Otherwise, the next default valence is considered. If no suitable default valence is found, zero is returned. The subvalence of an atom whose element has no default valences (i.e., no listing in Table~\ref{table:default-valences}) is zero. For example, an iron atom always has a subvalence of zero, regardless of its bonding.

@@

For a representative subvalence computation, consider an isolated nitrogen atom without virtual hydrogens. The atom's bond order sum is zero. Its default valences are 3 and 5. The difference between the atom's bond order sum and the first default valence is found to be three ($3 - 0$). Therefore, the subvalence of this atom is three.

@@

The phosphorous atom in phosphorous acid (H\textsubscript{3}PO\textsubscript{3}) illustrates the use of Algorithm \ref{algorithm:subvalence} for atoms having multiple target valences. Bond order sum is four ($2 + 1 + 1$). The first default valence is 3, but subtracting valence from that value yields a negative number ($3 - 4$). Continuing to the next default valence, five, a difference of 1 is obtained. Therefore, the subvalence of the phosphorous-bearing atom is reported as one.

@@

If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported. The meaning of such an atomic state may seem suspect because an atom without a free valence can not perform the required promotion of an elided bond during deselection. As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition. Returning zero avoids miscalculation of the implicit hydrogen count.

An atom's implicit hydrogen count may or may not correlate with chemical intuition or experimental data. Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}). We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2). However, the subvalence for the phosphorous atom is found to be three ($2 + 1$). The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two. To represent this atom and other like it, virtual hydrogens must be used.

@@

The same procedure works in reverse when encoding the \ttt{direction} attribute (Figure~\ref{fig:assigning-ppb}). 2-butene having the \textit{syn} conformation is modeled using a diagram whose left- and right-hand sides point downward and upward, respectively. The parent atom, a primary carbon, is identified. It is bonded to one neighbor through a PPB. To match the diagram, a direction of \ttt{Down} is assigned. The child atom is attached to its own child through a double bond. Finally, the grandchild is attached to its own child through a PPB whose direction is consistent with the diagram. A direction of \ttt{Up} is therefore assigned.

@@

An exception to the underspecification rule applies in the case of conjugated polyenes (Figure~\ref{fig:bond-error-state-exceptions}). Here, a PPB may run from parent to child without one running from grandchild to great-grandchild. No error is reported in such a case and the conformation of the terminal double bond remains undefined.

@@

Cyclic conjugated polyenes present special challenges to the use of PPBs, as exemplified by cyclooctatetraene (Figure~\ref{fig:cyclooctatetraene}). Although the (\textit{syn}, \textit{syn}, \textit{syn}, \textit{syn}) conformation is expressable, the (\textit{syn}, \textit{syn}, \textit{syn}, \textit{anti}) conformation is not. This problem arises because the conformation for two different double bonds is set by the same PPB.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{cyclooctatetraene.pdf}
    \caption{Cyclooctatetraene. The all-(\textit{syn}) conformation is expressable (left), but the (\textit{syn},\textit{syn},\textit{syn},\textit{anti}) conformation is not (right). Although the local conformations for the latter are all valid, the global conformation introduces a conflict, the direction marked in red.}
    \label{fig:cyclooctatetraene}
\end{figure}

@@

Balsa limits configuration to the special case of an atom with exactly four substituents placed at the vertices of a tetrahedron. Here, \enquote{substituent} means an atomic neighbor or virtual hydrogen. Lone electron pairs are not substituents.

The configuration about an atom can specified through its \ttt{parity} attribute. This attribute, together with a general convention, determine the relative three-dimensional positioning of substituents about a central atom (Figure~\ref{fig:atom-parity}). First, the parent of the central atom is identified. Then the frame of reference is shifted by sighting down the bond from the central atom's parent to the central atom. An equilateral triangle is then inscribed about the central atom. The bonds from the central atom are then iterated in order. If the parity of the central atom equals \ttt{Clockwise}, children placed around the triangle in clockwise order. Otherwise the children are placed in counterclockwise order.

@@

One additional modification applies to a central atom without a parent. The frame of reference is shifted by sighting along the bond from the first child to the central atom. The remaining children are then placed around the triangle in either counterclockwise or clockwise order as before. If the central atom bears a virtual hydrogen, it is considered to be the first child.

@@

It is sometimes useful to manipulate a configuration in a way that preserves the relative three-dimensional positioning of neighbor atoms. This process is called \textit{transformation}. Five operations suffice to transform any configuration into any other (Figure~\ref{fig:configurational-transformations}):

\begin{itemize}
    \item{Virtualize. Replaces an atomic first hydrogen child with a virtual hydrogen.}
    \item{Reify. Replaces a virtual hydrogen with an atomic hydrogen first child.}
    \item{Swap Children. Exchanges any two children while simultaneously toggling the configurational descriptor.}
    \item{Slide Down. Sets the parent node as the first child. Disabled if virtual hydrogen is present.}
    \item{Slide Up. Sets the parent of a parentless node as the first child. Disabled if virtual hydrogen is present.}
\end{itemize}
